---
title: "MCL_thesis_analysis"
author: "Heona"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    highlight: 
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
knitr:
  opts_chunk:
    echo: true
    message: false
    warning: false
    fig.path: "figures/"
    fig.width: 8
    fig.height: 6
    dpi: 300
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(janitor)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(cowplot)
library(randomcoloR)
library(ComplexHeatmap)
library(circlize)
library(tibble)
library(psych)
library(pcaMethods)
library(scrime)
library(tidyverse)
library(HarmonizR)
library(readxl)
library(limma)
library(sva)
library(stringr)
library(visdat)
library(DreamAI)
library(BayesDeBulk)
library(umap)
library(cola)
```

# MCL 1-100 Data Analysis

# RNA data 

## Import the data & merge
I have RNA1 and RNA2 from 2 different sources and have to unite them.
```{r rna data extraction}
## RNA data extraction

rna_1 <- readRDS("/Users/heona/git-repos/MCL_thesis_repo/data/raw_data/20250303_MCL_RNAseq1.rds")
rna_2 <- readRDS("/Users/heona/git-repos/MCL_thesis_repo/data/raw_data/20250303_MCL_RNAseq2.rds") 

print(paste("Missing values in rna_1:", sum(is.na(rna_1))))
print(paste("Missing values in rna_2:", sum(is.na(rna_2))))

rna_1_data <- rna_1 %>%
  as.data.frame() %>%
  dplyr::select(where(~ !any(is.na(.))))

rna_2_data <- rna_2 %>%
  as.data.frame() %>%
  dplyr::select(where(~ !any(is.na(.))))
```

First, I need to unify the column names.

dplyr::select(any_of(colnames(rna_data))) searches for overlapping column names, we still have different column names in each matrix - I have to change them to have uniform colnames.

```{r rna 1 and 2 mapping and unifying colnames}
mapping <- read_excel("/Users/heona/git-repos/MCL_thesis_repo/data/raw_data/20240830_MCL_IDs_Heona_V2_reduced.xlsx") 

#create a name vector, which replaces the RNAseq_ID names with the MS2_ID names
name_map <- setNames (mapping$MS2_ID, mapping$RNAseq_ID)

# rename columns in rna_2_data
colnames(rna_2_data) <- ifelse(colnames(rna_2_data) %in% names(name_map),
                               name_map[colnames(rna_2_data)], 
                               colnames(rna_2_data))

#create a name vector, which replaces the Cegat_id names with the MS2_ID names
name_map_2 <- setNames (mapping$MS2_ID, mapping$Cegat_id)

# only keep the columns that could be mapped, omit the missing ones that are still named "S..."
rna_2_data <- rna_2_data[, !grepl("^MCL", colnames(rna_2_data))]

# rename columns in rna_1_data
colnames(rna_1_data) <- ifelse(colnames(rna_1_data) %in% names(name_map_2),
                               name_map_2[colnames(rna_1_data)], 
                               colnames(rna_1_data))

# only keep the columns that could be mapped, omit the missing ones that are still named "S..."
rna_1_data <- rna_1_data[, !grepl("^S", colnames(rna_1_data))]
cat("dimension of rna_1_data:", dim(rna_1_data), "\n")

#hhow many gene names are overlapping
cat("shared rownames:", sum(rownames(rna_1) %in% rownames(rna_2)), "\n")
cat("rows only in rna_1:", sum(!(rownames(rna_1) %in% rownames(rna_2))), "\n")
cat("rows only in rna_2:",sum(!(rownames(rna_2) %in% rownames(rna_1))), "\n")

#to make a full join of rna 1 and 2, move rownames to a new column called "gene" - bc full_join can only work with columns, not rownames
rna_1_data$gene <- rownames(rna_1_data)
rna_2_data$gene <- rownames(rna_2_data)

# now full_join by gene
rna_data <- full_join(rna_1_data, rna_2_data, by = "gene")
rownames(rna_data) <- rna_data$gene # set the column gene back to rownames
rna_data$gene <- NULL # delete the column gene
```
rna_1_data and rna_2_data are from different labs and measured different genes.
There are 9995 overlapping genes and about 5000 genes unique to rna_1 and 2000 unique to rna_2.

## Merge with only shared genes

For further analysis, we need to proceed with only the data on the genes that were shared between both datasets.

```{r create rna_data_shared}
shared_genes <- intersect(rownames(rna_1_data), rownames(rna_2_data))

rna_data_shared <- cbind(rna_1_data[shared_genes, ], rna_2_data[shared_genes, ]) %>%
  select(-gene)
```

## PCA

```{r check for rna batch effect}

sum(is.na(rna_data_shared))

# 1. Filter out genes with too much missing data
threshold <- 0.9
rna_pca_df <- rna_data_shared %>%
  filter(rowMeans(is.na(.)) < threshold)

# Standard PCA (much faster)
rna_pca <- prcomp(t(as.matrix(rna_pca_df)), center = TRUE, scale. = TRUE)

# Extract scores
pca_out <- as.data.frame(rna_pca$x[, 1:2]) %>%
  rownames_to_column("Sample_ID")

# For variance explained, use:
var_explained <- summary(rna_pca)$importance[2, 1:2] * 100

# 5. for ex. "928_04" is plex_sample
pca_out <- pca_out %>%
  separate(Sample_ID, into = c("Plex", "Sample"), sep = "_", remove = FALSE)

# 6. Plot
ggplot(pca_out, aes(PC1, PC2, color = Plex)) +
  geom_point(size = 3) +
  theme_classic() +
  xlab(paste0("PC1 (", round(var_explained[1], 2), "%)")) +
  ylab(paste0("PC2 (", round(var_explained[2], 2), "%)")) +
  labs(title = "PCA of shared RNA-seq Samples")
```

The PCA shows a strong batch effect between the two datasets, and also within the plexes.
For BayesdeBulk, which we will perform with this rna data, it is crucial to not have any batch effects!


## Data check

```{r negative values and data range in rna data}
# Count how many values are below 0 and get the range of negative values
cat("=== rna_data_shared ===\n")
cat("Total sum:", sum(rna_data_shared), "\n")
cat("Data range:", range(rna_data_shared, na.rm = TRUE), "\n")
cat("Negative values:", sum(rna_data_shared < 0, na.rm = TRUE), "\n")
cat("Range of negative values:", range(rna_data_shared[rna_data_shared < 0], na.rm = TRUE), "\n\n")

cat("=== rna_1_data ===\n")
cat("Total sum:", sum(rna_1_data[ , sapply(rna_1_data, is.numeric)], na.rm = TRUE), "\n")
cat("Data range:", range(rna_1_data[ , sapply(rna_1_data, is.numeric)], na.rm = TRUE), "\n")
cat("Negative values:", sum(rna_1_data < 0, na.rm = TRUE), "\n")
cat("Range of negative values:", range(rna_1_data[rna_1_data < 0], na.rm = TRUE), "\n\n")

cat("=== rna_2_data ===\n")
cat("Total sum:", sum(rna_2_data[ , sapply(rna_2_data, is.numeric)], na.rm = TRUE), "\n")
cat("Data range:", range(rna_2_data[ , sapply(rna_2_data, is.numeric)], na.rm = TRUE), "\n")
cat("Negative values:", sum(rna_2_data < 0, na.rm = TRUE), "\n")
cat("Range of negative values:", range(rna_2_data[rna_2_data < 0], na.rm = TRUE), "\n")

```
Expression values ranged from approximately -2,8 to 16,2 in samples from lab 1 and from -3,5 to 18,3 in samples from lab 2.

### Normalization pattern

Since I have negative values in my RNA datasets that hinder batch correction methods to function, I want to check if my RNA data is normalized and that is the cause of the negative values.

```{r check normalization in rna}
library(tidyverse)
library(cowplot)

rna_data_shared %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  separate(sample, into = c("plex", "number"), sep = "_", remove = FALSE) %>%
  ggplot(aes(x = sample, y = expression, fill = plex)) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(angle = 90, size = 6, vjust = 0.5, hjust = 1)) +
  labs(x = "", y = "log2 (expression)") +
  scale_fill_brewer(palette = "Set3")

# Density plots
rna_data_shared %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  separate(sample, into = c("plex", "number"), sep = "_", remove = FALSE) %>%
  ggplot(aes(x = expression, color = plex)) +
  geom_density() +
  theme_classic() +
  labs(x = "log2 (expression)", color = "Plex") +
  scale_color_brewer(palette = "Set3")

rna_data_shared %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "expression") %>%
  separate(sample, into = c("plex", "number"), sep = "_", remove = FALSE) %>%
  ggplot(aes(x = expression, color = plex, fill = plex, group = sample)) +
  geom_density(fill = NA) +
  theme_classic() +
  labs(x = "log2 (expression)", color = "Plex", fill = "Plex") +
  scale_color_brewer(palette = "Set3") +
  scale_fill_brewer(palette = "Set3")

# Check column medians
rna_data_shared %>%
  summarise(across(everything(), ~median(.x, na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "sample", values_to = "median") %>%
  summary()
```

The boxplot visualization shows aligned medians within samples of the same lab origin, confirming normalization within each batch while revealing differences in normalization between the two laboratories


## Batch correction with ComBat

with lab origin as batch variable

The samples starting with 9 are from rna_1, the samples starting with 7 are from rna_2.

```{r combat on rna_data_shared}

batch_vector <- ifelse(grepl("^9", colnames(rna_data_shared)), "rna_1", "rna_2")
modcombat <- model.matrix(~1, data = data.frame(batch = batch_vector))

rna_combat <- ComBat(dat = as.matrix(rna_data_shared),
                        batch = batch_vector,
                        mod = modcombat,
                        par.prior = TRUE)
```

#### PCA after Batch correction

```{r standard PCA of rna after combat}
# Standard PCA
rna_pca_corrected <- prcomp(t(as.matrix(rna_combat)), scale. = TRUE)

# Create data frame for plotting
pca_out_corrected <- as.data.frame(rna_pca_corrected$x[, 1:2]) %>%
  rownames_to_column("Sample_ID") %>%
  separate(Sample_ID, into = c("Plex", "Sample"), sep = "_", remove = FALSE)

# Plot
ggplot(pca_out_corrected, aes(PC1, PC2, color = Plex)) +
  geom_point(size = 3) +
  theme_classic() +
  xlab(sprintf("PC1 (%.2f%%)", summary(rna_pca_corrected)$importance[2,1] * 100)) +
  ylab(sprintf("PC2 (%.2f%%)", summary(rna_pca_corrected)$importance[2,2] * 100)) +
  labs(title = "PCA of rna_combat (After ComBat)")
```

color by lab origin:
```{r standard PCA of rna after combat colored by lab origin}
# Standard PCA
pca_rna_combat <- prcomp(t(as.matrix(rna_combat)), scale. = TRUE)

# Create data frame for plotting
pca_df <- data.frame(
  PC1 = pca_rna_combat$x[,1],
  PC2 = pca_rna_combat$x[,2],
  Sample_ID = rownames(pca_rna_combat$x)
) %>%
  mutate(Lab = ifelse(substr(Sample_ID, 1, 1) == "9", "RNA_1", "RNA_2"))

# Plot
ggplot(pca_df, aes(PC1, PC2, color = Lab)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("RNA_1" = "#E41A1C", "RNA_2" = "#377EB8")) +
  theme_classic() +
  labs(
    x = sprintf("PC1 (%.1f%%)", summary(pca_rna_combat)$importance[2,1] * 100),
    y = sprintf("PC2 (%.1f%%)", summary(pca_rna_combat)$importance[2,2] * 100),
    title = "PCA of rna_combat (After ComBat) - colored by Lab"
  )
```

It looks like the batch correction was successful, the dominant separation by lab has been eliminated.


#Proteomic data

## Import the data

```{r data import}
getwd()
setwd("/Users/heona/git-repos/MCL_thesis_repo")
mcl_original <- read.table("/Users/heona/git-repos/MCL_thesis_repo/data/raw_data/MCL_1-100_Cologne_Essen.txt", header = TRUE, sep = "\t")%>%
  janitor::clean_names()
```

## Quick raw data check

```{r dim}
dim(mcl_original)
```

The dataset contains 6327 rows and 253 columns.
I select only the relevant columns and rename them.

```{r selecting and renaming data}
mcl_unfiltered <- mcl_original %>%
  dplyr::select(uniprot_i_ds, gene_names, contains("reporter_intensity_corrected")) %>%
  dplyr::rename("uniprot_id" = "uniprot_i_ds" ) %>%
  dplyr::rename("gene_id" = "gene_names") %>%
  dplyr::rename_with(.cols = -c("uniprot_id", "gene_id"), .fn = ~ stringr::str_remove(string = ., "reporter_intensity_corrected_"))

dim(mcl_unfiltered)
```

The transformed dataset mcl_unfiltered contains 6327 rows and 244 columns.

### Missing values

Check for number of missing values - R likes to work with NA as missing values, but in our dataset missing values are NaN.

```{r missing values}

print(paste("Total number of missing values: ", sum(is.na(mcl_unfiltered))))

print(paste("Proportion of missing values:", 
            round(sum(is.na(mcl_unfiltered)) / length(as.matrix(mcl_unfiltered)) * 100, 2), "%"))
```

### Initial visualization of NAs

```{r visualize NAs first for r1 and r2}

# Process both replicates together
missing_per_sample <- bind_rows(
  mcl_unfiltered %>%
    select(contains("_r1_")) %>%
    summarise(across(everything(), ~sum(is.na(.)) / n() * 100)) %>%
    pivot_longer(everything(), names_to = "sample", values_to = "pct_missing") %>%
    mutate(replicate = "r1"),
  
  mcl_unfiltered %>%
    select(contains("_r2_")) %>%
    summarise(across(everything(), ~sum(is.na(.)) / n() * 100)) %>%
    pivot_longer(everything(), names_to = "sample", values_to = "pct_missing") %>%
    mutate(replicate = "r2")
) %>%
  mutate(plex = str_extract(sample, "\\d{3}$"))

# Faceted plot
ggplot(missing_per_sample, aes(x = reorder(sample, as.numeric(plex)), y = pct_missing, fill = plex)) +
  geom_bar(stat = "identity") +
  facet_wrap(~replicate, scales = "free_x", ncol = 1) +
  labs(x = "Samples", 
       y = "% of proteins missing", 
       fill = "Plex",
       title = "Missing values per sample") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 4))
```

Samples across plexes show varying % of missing values, while within the plex usually have the same % of missingness. The spikes in missingness are the "empty" plexes, those that were labeled but not filled with samples.The mass spec does not measure perfectly, and slight overlap is usual, explaining the measurements of some signals even in the empty samples.
I exclude replicate 1 and 2 of the TMT-labels that were not used in the plexes.
Also, I exclude the plexes 949 ad 783 because they did not yield enough data (> 50% missing values)

```{r only select existing columns}

mcl <- mcl_unfiltered %>%
  select(-c("8_r1_775", "9_r1_775", "10_r1_775", # from Plex 5 from MCL1-54
                          "1_r1_930", "2_r1_930", "3_r1_930", "4_r1_930", "5_r1_930", # from Plex 3 from MCL56-100
                          "8_r2_775", "9_r2_775", "10_r2_775", # r2
                          "1_r2_930", "2_r2_930", "3_r2_930", "4_r2_930", "5_r2_930", # r2
                          )) %>%
  select(-contains("949")) %>%
  select(-contains("783"))
```
I convert NaN to NA and remove rows with only NA, then check the data again.

### Data on protein identified
```{r convert NaN and check data}
#convert NaN to NA
mcl[mcl == "NaN"] <- NA_integer_

#convert Zero values to NA
mcl[,-c(1:2)][mcl[,-c(1:2)] == "0"] <- NA_integer_

#remove rows with only NA
mcl <- mcl %>%
  filter(rowSums(is.na(mcl[,-c(1:2)])) != ncol(mcl[,-c(1:2)]))
```
Data check:
```{r print metrics for NA}
dim(mcl)
print(paste("Total number of proteins identified:", nrow(mcl)))

na_per_sample <- colSums(is.na(mcl[,-c(1:2)]))

print(paste("Median number of proteins identified per sample:", nrow(mcl) - median(na_per_sample)))
print(paste("Minimum number of proteins identified per sample:", nrow(mcl) - max(na_per_sample)))
print(paste("Maximum number of proteins identified per sample:", nrow(mcl) - min(na_per_sample)))

print(paste("Total number of missing values:", sum(is.na(mcl))))
print(paste("Proportion of missing values:", 
            round(sum(is.na(mcl)) / length(as.matrix(mcl)) * 100, 2), "%"))
print(paste("Negative values:", sum(mcl[sapply(mcl, is.numeric)] < 0, na.rm = TRUE)))
```
Proteomics typically has 20-40% missing values, which is much higher than RNA-seq. This is normal and expected due to the stochastic nature of mass spectrometry.


### Duplicates

check for duplicated features and samples:

```{r duplicates}
print(paste("Number of duplicated features: ", sum(duplicated(mcl$uniprot_id))))
print(paste("Number of duplicated samples: ", sum(duplicated(colnames(mcl[,-c(1:2)])))))
```

### Raw data distribution of mcl

inspect how the data is distributed without any correction

```{r raw data distribution}

order_vec <- colnames(mcl[,-c(1:2)])

# both plots without legend
p1 <- mcl %>%
  dplyr::select(-gene_id) %>%
  pivot_longer(!uniprot_id, names_to = "Prot_id", values_to = "intensity") %>%
  mutate(splitter = Prot_id) %>%
  separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
  filter(grepl("_r1_", Prot_id)) %>%
  ggplot(aes(factor(Prot_id, levels = order_vec), log2(intensity), fill = plex)) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(size = 5, angle = 90, hjust = 1, vjust = 1),
        legend.position = "none") +
  labs(x = "", y = "log2(intensity)", title = "Replicate 1") +
  scale_fill_brewer(palette = "Set3")

p2 <- mcl %>%
  dplyr::select(-gene_id) %>%
  pivot_longer(!uniprot_id, names_to = "Prot_id", values_to = "intensity") %>%
  mutate(splitter = Prot_id) %>%
  separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
  filter(grepl("_r2_", Prot_id)) %>%
  ggplot(aes(factor(Prot_id, levels = order_vec), log2(intensity), fill = plex)) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(size = 5, angle = 90, hjust = 1, vjust = 1),
        legend.position = "none") +
  labs(x = "", y = "log2(intensity)", title = "Replicate 2") +
  scale_fill_brewer(palette = "Set3")

# legend from one plot
legend <- get_legend(
  p1 + theme(legend.position = "right")
)

# combine plots and legend
plot_grid(
  plot_grid(p1, p2, nrow = 2),
  legend,
  rel_widths = c(1, 0.1)
)
```
This boxplot visualization of raw protein intensities on a log₂-scale reveals consistent intensity distributions between replicates while showing variations both within and across plexes indicating the need for normalization.


### Replicate correlation

Each plex was measured in two replicates - they should be similar - so we investigate the correlation between both replicates.

```{r Heatmap replicate correlation}

mcl_rep1 <- mcl %>%
  dplyr::select(contains("r1"))
mcl_rep2 <- mcl %>%
  dplyr::select(contains("r2"))

#set the color-vector 
color_fun_corr = colorRamp2(c(-1, 0, 1), c("blue", "white", "red"))

cor(mcl_rep1, mcl_rep2, method = "spearman", use = "pairwise.complete.obs") %>%
  Heatmap(as.matrix(.),
                   column_title = "Replicate 1",
                   row_title = "Replicate 2",
                   col = color_fun_corr,
                   show_row_names = FALSE,
                   show_column_names = FALSE,
                   cluster_rows = FALSE,
                   cluster_columns = FALSE,
                   name = "Spearman's R") %>%
  draw()
```
This shows an acceptable correlation, though optimal would be a higher contrast between the diagonal and off-diagonal correlations. Still, we can combine the replicate values to an average value.


### Combine replicates & reshape dataframe from wide to long

I combine the replicates to a dataframe with the average of r1 and r2.

```{r combine replicates}
mcl_r1 <- mcl %>%
  dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "samples_r1",
                        values_to = "count_R1")

mcl_r2 <- mcl %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "samples_r2",
                        values_to = "count_R2")

# mearge mcl_r1 and mcl_r2 and add an average column
mcl_long <- cbind(mcl_r1, mcl_r2[, -1]) # [, -1] to leave out the second uniprot_id column

mcl_avg <- mcl_long %>%
  rowwise() %>%
  mutate(avg_count = mean(c(count_R1, count_R2), na.rm = TRUE)) 

#reformat the NaN in the new column to NA
mcl_avg[mcl_avg == "NaN"] <- NA_integer_
```

### Visualize NAs

```{r visualize NAs of mcl_avg}
comb <- mcl_avg %>%
  group_by(samples_r1) %>%
  summarise(na_rate = sum(is.na(avg_count))) %>%
  mutate(samples_r2 = samples_r1) %>% 
  separate(samples_r2, c("remove", "Plex"), sep = "r1_") %>% 
  mutate(na_perc = na_rate / nrow(mcl)) 

ggplot(comb, aes(
    x = forcats::fct_reorder(samples_r1, na_perc),
    y = na_perc,
    fill = Plex
  )) +
  geom_col() +
  theme_cowplot() +
  labs(x = "samples", y = "[%] of proteins missing") +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line.y =  element_line(colour = "black"),
    axis.line.x =  element_line(colour = "black"),
    axis.text.y = element_text(colour = "black"),
    legend.title = element_blank(),
    axis.text.x = element_text(
      size = 6,
      angle = 90,
      hjust = 1,
      vjust = 1
    ),
    legend.position = "none"
  ) 
```


### Reference channel correlation

The reference channel is the 11th sample, so I only select the 11th sample of each plex as ref_cor.

Then, cor() calculates the pairwise Spearman correlations for the selected columns starting with 11, considering only complete observations (pairwise.complete.obs).
The resulting correlation matrix is converted to a data frame using as.data.frame().
The mutate() function adds new column "type" with the value "ref" to distinguish it as the reference-channel correlation.
pivot_longer() converts the correlation matrix from wide format to long format, resulting in a tibble with two columns: "Prot_id" and "count" (the correlation values) filter(count != 1) removes self-correlation values (correlation of a variable with itself), as these values are always 1 in correlation matrices.

```{r ref channel correlation}
mcl_wide <- mcl_avg %>%
  dplyr::select(c("uniprot_id", "samples_r1", "avg_count")) %>% #mcl_wide has the average count, we only take the name of samples_r1 (in case of confusion)
  pivot_wider(names_from = "samples_r1", values_from = avg_count)

#reference-channel correlation 
ref_cor11 <- mcl_wide %>%
  dplyr::select(starts_with("11_")) %>% 
  cor(., method = "spearman", use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  mutate(type = "ref") %>%
  pivot_longer(!c(type), names_to = "samples", values_to = "count") %>%
  filter(count != 1)

head(ref_cor11, n = 5)

ref_cor <- ref_cor11[duplicated(ref_cor11$count),] # remove duplicates that appear twice bc of pivot_longer

head(ref_cor, n = 5)

ggplot(ref_cor, aes(count)) +
  geom_density(fill = "#2a9d8f") +
  cowplot::theme_cowplot() +
  labs(x = "Spearman's R" ) 
```
The reference channel correlation ensures that measurements are accurate and not biased by technical irregularities. Spearman’s R showed an overall positive correlation for reference channels and a peak around 0.8, when ideally it would be at > 0.9. This sub-optimal correlation suggested the presence of technical variation between plexes, requiring batch effect correction.

## Sample loading normalization

To correct for differences in total protein amount loaded across samples and plexes, sample loading normalization is applied. A "target" is defined as the median total intensity across all samples (median of all column sums). For each sample, a normalization factor is calculated by dividing the target by the sample’s total intensity. 
Each protein intensity value is then multiplied by its corresponding normalization factor, scaling all samples to a common total intensity level while preserving the relative differences between individual proteins. To verify successful normalization, protein intensities are again visualized by boxplots on a log2-scale.

```{r sample loading normalization}
library(dplyr)

# create a vector of plexes
plex_vec <- unique(
  stringr::str_remove(
    colnames(mcl[,-c(1,2)]), 
    "._r1_|._r2_|.._r1_|.._r2_")
  ) 

# Store each plex into a list embedment
plex_list <- list() 

for(i in plex_vec) {
  plex_list[[i]] <- mcl %>% 
    dplyr::select(contains(i))
}

#create the target scaling factor 
colsum_vec <- c() 

for(i in plex_vec) {
  df <- plex_list[[i]]
  colsum_vec <- c(colsum_vec, colSums(df, na.rm = TRUE))
}

target <- median(colsum_vec, na.rm = TRUE)

#correct per plex with a helper function
run_sl_correct <- function(Y) { 
  norm_facs <- target / colSums(Y, na.rm = TRUE) # calculate normalization factor for each column in matrix Y
  output_sl <- sweep(Y, 2, norm_facs, FUN = "*") # multiplies norm_facs with each column
  return(output_sl) # output_sl is the normalized matrix
}

data_sl_repl <- plex_list%>%
  purrr::map(run_sl_correct) %>% 
  bind_cols()
```
visualize effect

```{r plot with one legend for thesis}
# Plots without legends
p1 <- data_sl_repl %>%
  tibble::rownames_to_column("n") %>%
  pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
  mutate(splitter = Prot_id) %>%
  separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
  filter(grepl("_r1_", Prot_id)) %>%
  ggplot(aes(
    factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
  )) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(
    size = 6,
    angle = 90,
    hjust = 1,
    vjust = 1
  ),
  legend.position = "none") +
  labs(x = "", y = "log2(intensity)", title = "Replicate 1") +
  scale_fill_brewer(palette = "Set3")

p2 <- data_sl_repl %>%
  tibble::rownames_to_column("n") %>%
  pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
  mutate(splitter = Prot_id) %>%
  separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
  filter(grepl("_r2_", Prot_id)) %>%
  ggplot(aes(
    factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
  )) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(
    size = 6,
    angle = 90,
    hjust = 1,
    vjust = 1
  ),
  legend.position = "none") +
  labs(x = "", y = "log2(intensity)", title = "Replicate 2") +
  scale_fill_brewer(palette = "Set3")

# Extract legend from one plot
legend <- get_legend(
  p1 + theme(legend.position = "right")
)

# combine plots and legend
plot_grid(
  plot_grid(p1, p2, nrow = 2),
  legend,
  rel_widths = c(1, 0.1)
)
```
After sample loading normalization protein intensities are effectively balanced across all samples. The boxplots show aligned medians and comparable intensity distributions across both replicates and plexes, indicating successful correction of technical variability in protein loading.

### Correlation between random sets of channels

I am comparing the ones with TMT-label 129C since they are present in every plex.

```{r corr between random sets after normalization}
library(psych)

sl_data <- data_sl_repl %>%
  dplyr::select(contains("r1")) %>%
  dplyr::select(contains(plex_vec[1:9])) %>%
  dplyr::select(contains("7_"))

pairs.panels(log2(sl_data), lm = TRUE, main = "Random channel 129C over plexes")
```

## Principal Component analysis (PPCA)

Let's further have a look into the general structure in the dataset by PCA.
Since we are working with a dataset that contains missing data, we probabilistic PCA (pPCA) that accounts for missing data <http://www.cs.columbia.edu/~blei/seminar/2020-representation/readings/TippingBishop1999.pdf>.

Highly correlated samples cluster together in a 2D graph The axes are ranked in order of importance - differences along x axis (PC1 = principle component 1) are more important than differences along the 2nd principal component on y-axis (PC2)

The scales: Large absolute values (e.g., -60 or 20) indicate that those samples differ significantly from the average sample along that principal component.
Smaller values (closer to 0) suggest that a sample is near the dataset’s mean in that dimension.

```{r initial pca}
library(pcaMethods)
library(scrime)

threshold <- 0.5

ppca_df <- data_sl_repl %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()

data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, c("Number", "Plex"), sep = "_r1_|_r2_")
ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_fill_brewer(palette = "Set3") +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)")) 
  
```

This is the PCA of the mcl data that was corrected with the normalization factor.
The PCA reveals prominent clustering of samples according to their plex membership, confirming that batch effects were a major source of variance in the dataset. 


## Internal reference scaling

The reference channel (channel 11, containing an identical pooled sample in each plex) serves as the basis for reference for protein-wise scaling within each plex.
For each protein, the geometric mean of intensities across all reference channels is defined as the reference average. A scaling factor is calculated by dividing the ref-erence average by the observed value in the reference channel of a plex. All values of that protein within the plex were then multiplied with the corresponding plex- and protein-specific scaling factor. 

```{r internal reference scaling}
#make a dataframe of the reference channels per plex 
irs_factors <- data_sl_repl %>% 
  dplyr::select(contains("11_"))

#calculate the geometric mean per sample as global reference for scaling
irs_factors$geomean <- apply(irs_factors, 1, function(x) exp(mean(log(x), na.rm = TRUE)))

#pull out sample names to identify  !!! Adjust removal position to geomeam column
irs_factor_vec = as.vector(names(irs_factors[,-c((length(plex_vec)*2)+1)]))
irs_factor_vec

#create the protein and reference channel wise scaling factor
for(i in irs_factor_vec) {
  factor <-  irs_factors$geomean / (irs_factors %>% dplyr::select(i))
  irs_factors[,paste0("fac_",i)] <- factor
}

irs_mult <- colnames(irs_factors %>% dplyr::select(contains("fac"))) %>% as.data.frame

irs_mult <- cbind(irs_mult, irs_mult) 

colnames(irs_mult) <- c("factor", "exp")

irs_mult<- irs_mult %>%
  separate(exp, c("discard", "plex"), sep = "\\_11_") %>%
  dplyr::select(-discard) %>%
  filter(plex !=  "r1_753") # i just took the first plex to start with

irs_mult_vec = as.vector(irs_mult$plex)

irs_fac_filt <- irs_factors %>% dplyr::select(contains("fac"))

all_irs <- (data_sl_repl %>% dplyr::select(contains( "r1_753"))) * unlist(irs_fac_filt %>% dplyr::select(contains( "r1_753"))) # starting with the first plex, the scaling is applied on all samples

for(i in irs_mult_vec) {
  all_irs <- cbind(all_irs, (data_sl_repl %>% dplyr::select(contains(i))) * unlist(irs_fac_filt %>% dplyr::select(contains(i))))
}
```
visualize effect

```{r boxplots after internal reference scaling}
# Create plots without legends
p1 <- all_irs %>%
  tibble::rownames_to_column("n") %>%
  pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
  mutate(splitter = Prot_id) %>%
  separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
  filter(grepl("_r1_", Prot_id)) %>%
  ggplot(aes(
    factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
  )) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(
    size = 6,
    angle = 90,
    hjust = 1,
    vjust = 1
  ),
  legend.position = "none") +
  labs(x = "", y = "log2(intensity)", title = "Replicate 1") +
  scale_fill_brewer(palette = "Set3")

p2 <- all_irs %>%
  tibble::rownames_to_column("n") %>%
  pivot_longer(!n, names_to = "Prot_id", values_to = "intensity") %>%
  mutate(splitter = Prot_id) %>%
  separate(splitter, c("remove", "plex"), sep = "_r1_|_r2_") %>%
  filter(grepl("_r2_", Prot_id)) %>%
  ggplot(aes(
    factor(Prot_id, levels = order_vec), log2(intensity), fill = plex
  )) +
  geom_boxplot() +
  theme_cowplot() +
  theme(axis.text.x = element_text(
    size = 6,
    angle = 90,
    hjust = 1,
    vjust = 1
  ),
  legend.position = "none") +
  labs(x = "", y = "log2(intensity)", title = "Replicate 2") +
  scale_fill_brewer(palette = "Set3")

# Extract legend from one plot
legend <- get_legend(
  p1 + theme(legend.position = "right")
)

# Combine plots and legend
plot_grid(
  plot_grid(p1, p2, nrow = 2),
  legend,
  rel_widths = c(1, 0.15)
)
```

### Correlation between random set of channels after IRS

```{r correlation between random set of channels 2}
sl_test_data <- all_irs %>%
  dplyr::select(contains("r1")) %>%
  dplyr::select(contains(plex_vec[1:9])) %>%
  dplyr::select(contains("7_"))

pairs.panels(log2(sl_test_data), lm = TRUE, main = "Random channel 129C over plexes")

```

### PPCA after IRS
```{r check ppca}
ppca_df <- all_irs %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()
data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)
ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, c("Number", "Plex"), sep = "_r1_|_r2_")
ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_fill_brewer(palette = "Set3") +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)"))

```

After internal reference scaling, data consistency was enhanced, improving reliability of downstream analyses. Correlation analysis of a random channel (channel 7) across plexes showed increased correlations (ρ ≈ 0.91–0.98) compared to pre-IRS values (ρ ≈ 0.63–0.80), confirming successful batch correction.
However, PPCA reveals that samples still clustered by plex, indicating residual batch effects requiring further correction. 

## Average replicates and log2-transform

I now average replicate values to a single value per protein per sample to create mcl_proteome_final. 

```{r merge replicates in smaller codes}
mcl_norm <- cbind(mcl[,c(1)], all_irs) %>%
  dplyr::select(-contains("11_"))

colnames(mcl_norm)[1] <- "uniprot_id"

# create the datasets for r1 and r2 first separately
mcl_long_norm_r1 <- mcl_norm %>%
    dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id1",
                        values_to = "count_R1")

mcl_long_norm_r2 <- mcl_norm %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "Prot_id2",
                        values_to = "count_R2") 

head(mcl_long_norm_r1, n = 3)
head(mcl_long_norm_r2, n = 3)

# now merge and leave out uniprot_id (first col) from r2 dataset

mcl_long_norm <- cbind(mcl_long_norm_r1, mcl_long_norm_r2[, -1]) 

mcl_long_norm <- mcl_long_norm %>%
  rowwise() %>%
  mutate(avg_count = mean(c(count_R1, count_R2), na.rm = TRUE))

#reformat the NA 
mcl_long_norm$avg_count[mcl_long_norm$avg_count == "NaN"] <- NA_integer_

mcl_wide_norm <- mcl_long_norm %>% 
  dplyr::select(uniprot_id, Prot_id1, avg_count) %>%
  pivot_wider(names_from = "Prot_id1", values_from = "avg_count")

```

Assign the gene_ids to the samples and log2-transform

```{r assign names to samples}
m_comb <- as.matrix(mcl_wide_norm[,-1])

rownames(m_comb) <- mcl_wide_norm$uniprot_id

# t() function swaps rows and cols

t_comb <- t(m_comb) %>% 
  as.data.frame() %>%
  rownames_to_column("Pre_id") %>%
  separate(Pre_id, c("number", "plex"), sep = "_r1_") %>%
  mutate(plex = paste0("P", as.character(plex))) %>%
  mutate(plex = paste0(as.character(plex), as.character(number))) %>%
  dplyr::select(-number) %>% 
  column_to_rownames("plex")

mcl_proteome <-
  t(t_comb) %>% 
  as.data.frame() %>% 
  rownames_to_column("uniprot_id")

uniprot_gene_mcl <- mcl %>% dplyr::select(uniprot_id, gene_id) 

mcl_proteome_final <-  left_join(uniprot_gene_mcl, mcl_proteome, by  = "uniprot_id") 

mcl_proteome_final[,3:ncol(mcl_proteome_final)]<- sapply(mcl_proteome_final[,3:ncol(mcl_proteome_final)], as.numeric)

mcl_proteome_final[,3:ncol(mcl_proteome_final)] <- log2(mcl_proteome_final[,3:ncol(mcl_proteome_final)]) # log-transformed!

```
### PPCA before HarmonizR

```{r mcl_proteome_final ppca}
ppca_df <- mcl_proteome_final[,-c(1,2)] %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()

data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  mutate(Plex = str_remove(str_sub(Prot_id, 1, 4), "P"))
# sep = 4 means separate at the 4th string character

ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_fill_brewer(palette = "Set3") +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)"))
```
missingness distribution
```{r final data NA distribution 1}
comb_df <- comb %>%
  dplyr::select(Plex, na_perc) %>%
  group_by(Plex) %>%
  summarise(na = mean(na_perc)) %>%
  mutate(Plex = paste0("P", Plex, sep = ""))

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, c("Plex", "Number"), sep = 4) %>%
  left_join(comb_df, 
            by = "Plex")

ggplot(ppca_out, aes(PC1, PC2, col = na)) +
  geom_point(size = 3) +
  scale_color_viridis_c()+
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)")) +
  labs(color = "% missing")
```

## HarmonizR 

with Plex = Batch, and with specifying parameters

HarmonizR integrates limma’s removeBatchEffect() function employing linear regres-sion and the batch effect correction method ComBat which uses an empirical Bayes framework. Unlike most harmonization tools that require complete datasets, HarmonizR can handle missing values by matrix dissection, without imputing or reducing data, which makes it suitable for proteomic data. 

I use the normalized log-transformed data mcl_proteome_final for HarmonizR.

```{r check missingness before HarmonizR}
print(paste("Total number of missing values: ", sum(is.na(mcl_proteome_final))))
prod(dim(mcl_proteome_final))
print(sum(is.na(mcl_proteome_final)) / prod(dim(mcl_proteome_final)))
```
We have 21,57% missing values before applying harmonizr.

```{r harmonizr specifying the correction parameters}
# 1: Create the batch vector information
# Extract sample names (excluding uniprot_id and gene_name columns)
# make batch_df with cols: samplename, numbered ID and batch
batch_df <- tibble(samplename = colnames(mcl_proteome_final[,-c(1,2)])) %>%
  rowid_to_column("ID") %>%
  relocate(ID, .after = samplename) %>%
  mutate(
    plex_number = sub("P(.{3}).*", "\\1", samplename),
    batch = case_when(
      plex_number == "753" ~ 1,
      plex_number == "757" ~ 2,
      plex_number == "764" ~ 3,
      plex_number == "772" ~ 4,
      plex_number == "775" ~ 5,
      plex_number == "920" ~ 6,
      plex_number == "928" ~ 7,
      plex_number == "930" ~ 8,
      plex_number == "935" ~ 9,
      TRUE ~ NA_real_
    )
  ) %>%
  select(samplename, ID, batch)

# Check batch assignments
table(batch_df$batch)

# 2: Prepare dataframe for harmonization
# Remove gene_name column and set uniprot_id as rownames
harmonize_df <- mcl_proteome_final[,-2] %>%
  column_to_rownames("uniprot_id")

# 3: Run HarmonizR with explicit parameters
mcl_harmonized <- harmonizR(
  data_as_input = harmonize_df,
  description_as_input = batch_df,
  algorithm = "ComBat",
  ComBat_mode = 3,  # parametric adjustment (1 = non-parametric if needed)
  plot = "samplemeans"  # generates diagnostic plots
)

# 4: Convert back to dataframe with protein IDs
mcl_proteome_harmonized <- mcl_harmonized %>%
  as.data.frame() %>%
  rownames_to_column("uniprot_id") %>%
  left_join(mcl_proteome_final[, c("uniprot_id", "gene_id")], by = "uniprot_id") %>%
  relocate(gene_id, .after = uniprot_id)

# store files in the processed_data folder

file.rename("cured_data.tsv", "data/processed_data/mcl_harmonized.tsv")
file.rename("cured_data.pdf", "data/processed_data/mcl_harmonized_diagnostics.pdf")
```

###PPCA after HarmonizR

```{r ppca mcl_proteome_harmonized}
ppca_df <- mcl_proteome_harmonized[,-c(1,2)] %>% 
  filter(rowMeans(is.na(.)) < threshold) %>%
  #feature-wise scaling and centering 
  rowScales() %>%
  as.data.frame()

data_ppca <- pcaMethods::pca(t(as.matrix(ppca_df)), method = "ppca", nPcs = 2, seed = 123)

ppca_out <- as.data.frame(scores(data_ppca)) %>%
  rownames_to_column("Prot_id") %>%
  separate(Prot_id, into = c("Plex", "Number"), sep = 4)
# sep = 4 means separate at the 4th string character

ggplot(ppca_out, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_fill_brewer(palette = "Set3") +
  theme_classic() +
  xlab(paste("PC1 (", round(data_ppca@R2[1] * 100, digits = 2), "%)")) +
  ylab(paste("PC2 (", round(data_ppca@R2[2] * 100, digits = 2), "%)"))
```

HarmonizR successfully corrected the batch effects.
We proceed with mcl_proteome_harmonized for our missing value imputation.

## Mapping and unifying colnames

```{r prot data unify colnames}
# for protein data, create a name vector, which replaces the MS_Label names with the MS2_ID names
name_map_ms <- setNames (mapping$MS2_ID, mapping$MS_Label)

proteome_data_log <- mcl_proteome_harmonized

# rename columns in mcl_proteome_cured_data to "753_01" to fit the same format -> prot_data
colnames(proteome_data_log) <- ifelse(colnames(proteome_data_log) %in% names(name_map_ms),
                               name_map_ms[colnames(proteome_data_log)], 
                               colnames(proteome_data_log))

```

proteome_data_log now has the same colnames as the rna_data_shared

We still have multiple genes in some gene_id rownames, that are seperated by ";" So we select one of the genes (the first) and drop the rest.
If the first of the row is already taken prior, the next is selected as unique identifier.


### Create unique gene_id identifiers

```{r unique gene_id identifiers - log transformed data}
# A vector to keep track of the genes that have already been used
used_genes <- c()

# A vector to store the new, unique gene IDs
unique_gene_id <- sapply(proteome_data_log$gene_id, function(genes_str) {
  
  # Split the string of genes by the semicolon
  candidate_genes <- str_split(genes_str, ";")[[1]]
  
  # Find the first gene in the list that has not been used yet
  for (gene in candidate_genes) {
    if (!gene %in% used_genes) {
      used_genes <<- c(used_genes, gene) # If found, add it to our list of used genes
      return(gene) # Return this gene as the chosen one for this row
    }
  }
  # If no unique gene is found for this row (all candidates were already used),
  # return NA as a placeholder.
  return(NA_character_)
}) 

# Add the new column of unique gene IDs to dataframe
proteome_data_log$unique_gene_id <- unique_gene_id

prot_data_log <- proteome_data_log %>%
  select(uniprot_id, gene_id, unique_gene_id, everything())

sum(is.na(unique_gene_id))
```

165 rows cannot be matched with a unique_gene_id, and thus have to be removed. We only work with those that have both gene_id and uniprot_id because we focus on the features that are gene/protein-code.

```{r prot_data only select columns unique_gene_id and samples - log transformed data}
prot_data_final_log <- prot_data_log %>%
  select(-uniprot_id, -gene_id) %>%
  filter(!is.na(unique_gene_id)) 

rownames(prot_data_final_log) <- prot_data_final_log$unique_gene_id
prot_data_final_log <- prot_data_final_log[,-1]
```

## DreamAI 

In order to apply BayesDeBulk deconvolution analysis with RNA and protein data, we need our datasets to be free of missing values.
To impute the missing values I use DreamAI.

### Missing values
```{r missing values exploration - mcl_proteome_harmonized}

na_percentage <- rowMeans(is.na(prot_data_final_log)) * 100 #
rows_above_50_na <- which(na_percentage > 50)

row_lowest_na <- which.min(na_percentage) 
min_na_percentage <- na_percentage[row_lowest_na] 

row_highest_na <- which.max(na_percentage) 
max_na_percentage <- na_percentage[row_highest_na] 

# Proportion of rows with >50% missing values
proportion_above_50_na <- sum(na_percentage > 50) / nrow(mcl_proteome_harmonized) *100

# Output results
list(
  num_rows_above_50_na = sum(na_percentage > 50),
  row_lowest_na = row_lowest_na,
  min_na_percentage = min_na_percentage,
  row_highest_na = row_highest_na,
  max_na_percentage = max_na_percentage,
  proportion_above_50_na = proportion_above_50_na
)
```

### Filtering for DreamAI
As a preparation, we need to filter the rows that have NA > 50%, 
We only take the columns that are also found in RNA data. For some, RNA data was missing and we have to exclude these.

```{r refine prot_data_final - log transformed data}
sum(is.na(prot_data_final_log))

# for DreamAI limit the threshold of missing values to 0.5
threshold <- 0.5

prot_data_final_dai_log <- prot_data_final_log %>%
  filter(rowMeans(is.na(.[,])) < threshold) %>%
  select(, any_of(colnames(rna_combat))) # only the columns which are found in rna_data. bc for some, rna data was missing and we have to exclude these

sum(is.na(prot_data_final_dai_log))

dim(prot_data_final_log)
dim(prot_data_final_dai_log)
dim(rna_combat)
```
The dimensions are now matched between RNA and protein data.

### Missing values after filtering

Before I run DreamAI, I want to see, how many missing values there are.

```{r missing values right before dreamai - log trasnformed data}

vis_miss(prot_data_final_dai_log[, order(colnames(prot_data_final_dai_log))]) +
  theme(
    axis.text.x = element_text(size = 4, angle = 90, vjust = 1, hjust = 1),
    plot.margin = ggplot2::margin(t = 50, r = 5, b = 5, l = 5)
  )
```

The graph shows how many missing values are in each sample now, it varies from 3% to 22%, on average 8,4%.
You can derive from this graph, that the missing values are not missing at random, they follow a pattern.

### Run DreamAI() 

on prot_data_final_dai_log

Install DreamAI:

```{r dream ai requirements, echo=TRUE, results='hide'}
require("cluster")
require("survival")
require("randomForest")
require("missForest")
require("glmnet")
require("Rcpp")
require("foreach")
require("itertools")
require("iterators")
require("Matrix")
require("devtools")
if (!requireNamespace("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("impute") #original code: BiocManager::install("impute", version = "3.8"), caused problems because my R version is newer (4.4) and is only compatible with BiocManager 3.19. If i need the 3.8 version i need to install R 3.5.x
require("impute")
BiocManager::version()

```
Run DreamAI
```{r run Dream AI on prot_data_final_dai_log , echo=TRUE}

library(DreamAI)

imputed_prot_data_log <- DreamAI(prot_data_final_dai_log, k = 10, maxiter_MF = 10, ntree = 100,
  maxnodes = NULL, maxiter_ADMIN = 30, tol = 10^(-2),
  gamma_ADMIN = NA, gamma = 50, CV = FALSE,
  fillmethod = "row_mean", maxiter_RegImpute = 10,
  conv_nrmse = 1e-06, iter_SpectroFM = 40, method = c("KNN",
  "MissForest", "ADMIN", "Birnn", "SpectroFM", "RegImpute"),
  out = c("Ensemble"))

#imputed_prot_data_log$Ensemble has the new data

#check NAs
sum(is.na(imputed_prot_data_log$Ensemble))

#make it a dataframe (and later maybe a matrix)
prot_data_dreamai_log <- imputed_prot_data_log$Ensemble %>%
  as.data.frame()
```
prot_data_dreamai_log is the imputed dataframe to further work with.

```{r reorder cols ascending}
order_asc = sort(colnames(prot_data_dreamai_log))

prot_data_dreamai_log <- prot_data_dreamai_log[, order_asc]
```

### PCA after DreamAI

of prot_data_dreamai_log

as a last check to not have batch effects
standard PCA (bc we have a complete dataset after DreamAI)

```{r standard PCA of prot_data_dreamai_log}
# Prepare data with scaling
pca_df_dreamai_log <- prot_data_dreamai_log %>%
  as.data.frame() %>%
  rowScales() %>%
  as.data.frame()

# Run standard PCA using prcomp (base R)
data_pca_dreamai_log <- prcomp(t(as.matrix(pca_df_dreamai_log)), 
                                center = FALSE,  # already centered via rowScales
                                scale. = FALSE)  # already scaled via rowScales

# Extract scores and separate sample IDs
pca_out_dreamai <- as.data.frame(data_pca_dreamai_log$x[, 1:2]) %>%
  rownames_to_column("Sample_id") %>%
  separate(Sample_id, into = c("Plex", "Number"), sep = "_")

# Calculate variance explained
var_explained <- summary(data_pca_dreamai_log)$importance[2, 1:2] * 100

# Plot
ggplot(pca_out_dreamai, aes(PC1, PC2, col = Plex)) +
  geom_point(size = 3) +
  scale_color_brewer(palette = "Set3") +
  theme_classic() +
  xlab(paste0("PC1 (", round(var_explained[1], 2), "%)")) +
  ylab(paste0("PC2 (", round(var_explained[2], 2), "%)")) +
  ggtitle("PCA of prot_data_dreamai_log")
```
No batch effect visible - which means we can take prot_data_dreamai_log for BayesdeBulk.


##BayesDeBulk

https://github.com/WangLab-MSSM/BayesDeBulk.git

using prot_data_dreamai_log and rna_combat

BayesDeBulk expects MATRICES with:
- Genes as ROWS
- Samples as COLUMNS

```{r checking prot_data_dreamai_log and rna_combat}
# Check your data structure
class(rna_combat)     
class(prot_data_dreamai_log)      

dim(rna_combat)        
dim(prot_data_dreamai_log)        

head(rownames(prot_data_dreamai_log)) # Should be gene names
head(colnames(prot_data_dreamai_log)) # Should be sample names

head(rownames(rna_combat))   # Should be gene names
head(colnames(rna_combat))   # Should be sample names
```

run BayesDeBulk

```{r bayesdb with prot_data_dreamai_log, warning=FALSE, message=FALSE, cache = TRUE}
out <-BayesDeBulk(n.iter=5000, burn.in=2500,
                  Y=list(rna_combat,prot_data_dreamai_log), 
                  markers = LM22_markers(list(rna_combat,prot_data_dreamai_log)))

bayesdb_out_log <- out$cell.fraction %>%
  as.data.frame()

write.csv(bayesdb_out_log, 
          file = "data/processed_data/bayesdb_out_log.csv", 
          row.names = TRUE) 
```



#### Cell composition data check

```{r bayesdb_out_log data check}
library(tidyverse)

# 1. Check structure
dim(bayesdb_out_log)
colnames(bayesdb_out_log)

# 2. Summary statistics per cell type
celltype_summary <- bayesdb_out_log %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "cell_type", values_to = "proportion") %>%
  group_by(cell_type) %>%
  summarise(
    median = median(proportion, na.rm = TRUE),
    mean = mean(proportion, na.rm = TRUE),
    sd = sd(proportion, na.rm = TRUE),
    min = min(proportion, na.rm = TRUE),
    max = max(proportion, na.rm = TRUE),
    cv = sd / mean * 100
  ) %>%
  arrange(desc(median))

print(celltype_summary, n = Inf)

# 3. Boxplot of cell type proportions
bayesdb_out_log %>%
  rownames_to_column("sample") %>%
  pivot_longer(-sample, names_to = "cell_type", values_to = "proportion") %>%
  ggplot(aes(x = reorder(cell_type, proportion, FUN = median), y = proportion)) +
  geom_boxplot() +
  coord_flip() +
  theme_classic() +
  labs(x = "Cell type", y = "Estimated proportion",
       title = "Cell type proportions across samples")

# 4. Heatmap
library(ComplexHeatmap)

Heatmap(
  t(as.matrix(bayesdb_out_log)),  # transpose so cell types are rows
  name = "Proportion",
  row_title = "Cell types",
  column_title = "Samples",
  show_column_names = FALSE,
  cluster_rows = TRUE,
  cluster_columns = TRUE
)

# 5. Variability plot (coefficient of variation)
celltype_summary %>%
  ggplot(aes(x = reorder(cell_type, cv), y = cv)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_classic() +
  labs(x = "Cell type", y = "Coefficient of variation (%)",
       title = "Variability of cell type proportions")

# 6. Quick summary for your thesis
cat("Number of cell types:", ncol(bayesdb_out_log), "\n")
cat("Number of samples:", nrow(bayesdb_out_log), "\n\n")

cat("Highest median abundance:\n")
print(head(celltype_summary, 3))

cat("\nLowest median abundance:\n")
print(tail(celltype_summary, 3))

cat("\nMost variable cell types (highest CV):\n")
print(celltype_summary %>% arrange(desc(cv)) %>% head(3))
```
#### UMAP for plexes

```{r umap for plexes}
umap_bdb_log <- umap(bayesdb_out_log)

umap_df <- data.frame(
  UMAP1 = umap_bdb_log$layout[,1],
  UMAP2 = umap_bdb_log$layout[,2],
  Sample = rownames(bayesdb_out_log)
) %>%
  separate(Sample, into = c("Plex", "Number"), sep = "_", remove = FALSE)

ggplot(umap_df, aes(UMAP1, UMAP2, color = Plex)) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set3") +
  theme_classic() +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "UMAP of bayesdb_out_log colored by Plex"
  )
```


#### UMAP for single plexes

```{r umap single plexes}

umap_bdb_log <- umap(bayesdb_out_log, n_neighbors = 15)

# Create data frame for plotting
umap_df <- data.frame(
  UMAP1 = umap_bdb_log$layout[,1],
  UMAP2 = umap_bdb_log$layout[,2],
  Sample = rownames(bayesdb_out_log)
) %>%
  separate(Sample, into = c("Plex", "Number"), sep = "_", remove = FALSE) %>%
  mutate(Highlight = ifelse(Plex %in% c("930", "935"), Plex, "Other"))

# Plot
ggplot(umap_df, aes(UMAP1, UMAP2, color = Highlight)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("930" = "#E41A1C", "935" = "#377EB8", "Other" = "grey80")) +
  theme_classic() +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "UMAP of bayesdb_out_log colored by Plex 930 and 935"
  )
```

UMAP of 928:
```{r UMAP of 928}
umap_bdb_log <- umap(bayesdb_out_log)

umap_df <- data.frame(
  UMAP1 = umap_bdb_log$layout[,1],
  UMAP2 = umap_bdb_log$layout[,2],
  Sample = rownames(bayesdb_out_log)
) %>%
  separate(Sample, into = c("Plex", "Number"), sep = "_", remove = FALSE) %>%
  mutate(Highlight = ifelse(Plex %in% c("928"), Plex, "Other"))

ggplot(umap_df, aes(UMAP1, UMAP2, color = Highlight)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("928" = "#E41A1C", "Other" = "grey80")) +
  theme_classic() +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "UMAP of bayesdb_out_log colored by Plex 928"
  )
```
UMAP of 775
```{r umap of 775}
umap_bdb_log <- umap(bayesdb_out_log)

umap_df <- data.frame(
  UMAP1 = umap_bdb_log$layout[,1],
  UMAP2 = umap_bdb_log$layout[,2],
  Sample = rownames(bayesdb_out_log)
) %>%
  separate(Sample, into = c("Plex", "Number"), sep = "_", remove = FALSE) %>%
  mutate(Highlight = ifelse(Plex %in% c("775"), Plex, "Other"))

ggplot(umap_df, aes(UMAP1, UMAP2, color = Highlight)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("775" = "#E41A1C", "Other" = "grey80")) +
  theme_classic() +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "UMAP of bayesdb_out_log colored by Plex 775"
  )
```

There are no big batch effects visible.

#### UMAP lab origin
```{r umap lab origin}
umap_df <- data.frame(
  UMAP1 = umap_bdb_log$layout[,1],
  UMAP2 = umap_bdb_log$layout[,2],
  Sample = rownames(bayesdb_out_log)
) %>%
  mutate(Lab = ifelse(substr(Sample, 1, 1) == "9", "RNA_1", "RNA_2"))

ggplot(umap_df, aes(UMAP1, UMAP2, color = Lab)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("RNA_1" = "#E41A1C", "RNA_2" = "#377EB8")) +
  theme_classic() +
  labs(
    x = "UMAP1",
    y = "UMAP2",
    title = "UMAP of bayesdb_out_log - colored by RNA source lab"
  )
```

#### Scaled UMAPs

```{r scaled bayesdb umaps}
# Scaled UMAP
set.seed(123)

bayesdb_log_scaled <- scale(bayesdb_out_log)
umap_scaled <- umap(bayesdb_log_scaled)

# Base data frame
umap_base <- data.frame(
  UMAP1 = umap_scaled$layout[,1],
  UMAP2 = umap_scaled$layout[,2],
  Sample = rownames(bayesdb_log_scaled)
) %>%
  separate(Sample, into = c("Plex", "Number"), sep = "_", remove = FALSE) %>%
  mutate(
    Lab = ifelse(substr(Sample, 1, 1) == "9", "RNA_1", "RNA_2"),
    Highlight_930_935 = case_when(
      Plex == "930" ~ "930",
      Plex == "935" ~ "935",
      TRUE ~ "Other"
    ),
    Highlight_928 = ifelse(Plex == "928", "928", "Other"),
    Highlight_775 = ifelse(Plex == "775", "775", "Other")
  )

# 1. Plex 930 and 935
p1 <- ggplot(umap_base, aes(UMAP1, UMAP2, color = Highlight_930_935)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("930" = "#E41A1C", "935" = "#377EB8", "Other" = "grey80"), name = "Plex") +
  theme_classic() +
  labs(title = "Plex 930 & 935")

# 2. Plex 928
p2 <- ggplot(umap_base, aes(UMAP1, UMAP2, color = Highlight_928)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("928" = "#E41A1C", "Other" = "grey80"), name = "Plex") +
  theme_classic() +
  labs(title = "Plex 928")

# 3. Plex 775
p3 <- ggplot(umap_base, aes(UMAP1, UMAP2, color = Highlight_775)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("775" = "#E41A1C", "Other" = "grey80"), name = "Plex") +
  theme_classic() +
  labs(title = "Plex 775")

# 4. All plexes
p4 <- ggplot(umap_base, aes(UMAP1, UMAP2, color = Plex)) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set3") +
  theme_classic() +
  labs(title = "All Plexes")

# 5. RNA source lab
p5 <- ggplot(umap_base, aes(UMAP1, UMAP2, color = Lab)) +
  geom_point(size = 2) +
  scale_color_manual(values = c("RNA_1" = "#E41A1C", "RNA_2" = "#377EB8")) +
  theme_classic() +
  labs(title = "RNA Source Lab")

# Combine all
library(patchwork)
(p1 + p2) / p3 +
  plot_annotation(title = "UMAP of bayesdb_out_log (scaled)", theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))

(p4 + p5) +
  plot_annotation(title = "UMAP of bayesdb_out_log (scaled)", theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))

(p1 + p2) / (p4 + p5) +
  plot_annotation(title = "UMAP of bayesdb_out_log (scaled)", theme = theme(plot.title = element_text(hjust = 0.5, face = "bold")))
```

## Cola clustering
on bayesdb_out_log

```{r cola on bayesdb_out_log}
library(cola)
library(dplyr)

graphics.off()

bayesdb_mat_log <- bayesdb_out_log %>%
  as.matrix() %>%
  t()  

# set seed for reproducibility!
set.seed(123) 

# to prevent the error message "#> Error in dev.off(i2): cannot shut down device 1 (the null device)"
while (!is.null(dev.list()))  dev.off()

#run cola
rl_bayesdb_log <- run_all_consensus_partition_methods(
  bayesdb_mat_log,
  top_value_method = c("SD", "MAD", "ATC"), 
  partition_method = c("hclust", "kmeans", "skmeans"),
  max_k = 6, 
  scale_rows = TRUE,
  cores = 4
)
# Generate full report
cola_report(rl_bayesdb_log, output_dir = "data/processed_data/cola_bayesdb_log", cores = 4)

best_results_bdb_log <- suggest_best_k(rl_bayesdb_log)
head(best_results_bdb_log, 5)

```


