---
title: "MCL_thesis_analysis"
author: "Heona"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    toc_depth: 2
  html_document:
    highlight: 
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
knitr:
  opts_chunk:
    echo: true
    message: false
    warning: false
    fig.path: "figures/"
    fig.width: 8
    fig.height: 6
    dpi: 300
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(janitor)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(cowplot)
library(randomcoloR)
library(ComplexHeatmap)
library(circlize)
library(tibble)
library(psych)
library(pcaMethods)
library(scrime)
library(tidyverse)
library(HarmonizR)
library(readxl)
library(limma)
library(sva)
library(stringr)
```

# MCL 1-100 Data Analysis

## Import the data

```{r data import}
getwd()
setwd("/Users/heona/git-repos/MCL_thesis_repo")
mcl_original <- read.table("/Users/heona/git-repos/MCL_thesis_repo/data/raw_data/MCL_1-100_Cologne_Essen.txt", header = TRUE, sep = "\t")%>%
  janitor::clean_names()
```

## Quick raw data check

```{r dim}
dim(mcl_original)
```

The dataset contains 6327 rows and 253 columns.
I select only the relevant columns and rename them.

```{r selecting and renaming data}
mcl_unfiltered <- mcl_original %>%
  dplyr::select(uniprot_i_ds, gene_names, contains("reporter_intensity_corrected")) %>%
  dplyr::rename("uniprot_id" = "uniprot_i_ds" ) %>%
  dplyr::rename("gene_id" = "gene_names") %>%
  dplyr::rename_with(.cols = -c("uniprot_id", "gene_id"), .fn = ~ stringr::str_remove(string = ., "reporter_intensity_corrected_"))

dim(mcl_unfiltered)
```

The transformed dataset mcl_unfiltered contains 6327 rows and 244 columns.

### Missing values

Check for number of missing values - R likes to work with NA as missing values, but in our dataset missing values are NaN.

```{r missing values}

print(paste("Total number of missing values: ", sum(is.na(mcl_unfiltered))))

print(paste("Proportion of missing values:", 
            round(sum(is.na(mcl_unfiltered)) / length(as.matrix(mcl_unfiltered)) * 100, 2), "%"))
```

### Initial visualiztaion of NAs

```{r visualize NAs first for r1 and r2}
library(tidyverse)

# Process both replicates together
missing_per_sample <- bind_rows(
  mcl_unfiltered %>%
    select(contains("_r1_")) %>%
    summarise(across(everything(), ~sum(is.na(.)) / n() * 100)) %>%
    pivot_longer(everything(), names_to = "sample", values_to = "pct_missing") %>%
    mutate(replicate = "r1"),
  
  mcl_unfiltered %>%
    select(contains("_r2_")) %>%
    summarise(across(everything(), ~sum(is.na(.)) / n() * 100)) %>%
    pivot_longer(everything(), names_to = "sample", values_to = "pct_missing") %>%
    mutate(replicate = "r2")
) %>%
  mutate(plex = str_extract(sample, "\\d{3}$"))

# Faceted plot
ggplot(missing_per_sample, aes(x = reorder(sample, as.numeric(plex)), y = pct_missing, fill = plex)) +
  geom_bar(stat = "identity") +
  facet_wrap(~replicate, scales = "free_x", ncol = 1) +
  labs(x = "Samples", 
       y = "% of proteins missing", 
       fill = "Plex",
       title = "Missing values per sample") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 4))
```

Samples across plexes show varying % of missing values, while within the plex usually have the same % of missingness. The spikes in missingness are the "empty" plexes, those that were labeled but not filled with samples.The mass spec does not measure perfectly, and slight overlap is usual, explaining the measurements of some signals even in the empty samples.
I exclude replicate 1 and 2 of the TMT-labels that were not used in the plexes.
Also, I exclude the plexes 949 ad 783 because they did not yield enough data (> 50% missing values)

```{r only select existing columns}

mcl <- mcl_unfiltered %>%
  select(-c("8_r1_775", "9_r1_775", "10_r1_775", # from Plex 5 from MCL1-54
                          "1_r1_930", "2_r1_930", "3_r1_930", "4_r1_930", "5_r1_930", # from Plex 3 from MCL56-100
                          "8_r2_775", "9_r2_775", "10_r2_775", # r2
                          "1_r2_930", "2_r2_930", "3_r2_930", "4_r2_930", "5_r2_930", # r2
                          )) %>%
  select(-contains("949")) %>%
  select(-contains("783"))
```
I convert NaN to NA and remove rows with only NA, then check the data again.

```{r convert NaN and check data}
#convert NaN to NA
mcl[mcl == "NaN"] <- NA_integer_

#convert Zero values to NA
mcl[,-c(1:2)][mcl[,-c(1:2)] == "0"] <- NA_integer_

#remove rows with only NA
mcl <- mcl %>%
  filter(rowSums(is.na(mcl[,-c(1:2)])) != ncol(mcl[,-c(1:2)]))

#data check
dim(mcl)
print(paste("Total number of missing values: ", sum(is.na(mcl))))
print(paste("Proportion of missing values:", 
            round(sum(is.na(mcl)) / length(as.matrix(mcl)) * 100, 2), "%"))
print(paste("Negative values:", sum(mcl[sapply(mcl, is.numeric)] < 0, na.rm = TRUE)))
```
Proteomics typically has 20-40% missing values, which is much higher than RNA-seq. This is normal and expected due to the stochastic nature of mass spectrometry.


### Duplicates

check for duplicated features and samples:

```{r duplicates}
print(paste("Number of duplicated features: ", sum(duplicated(mcl$uniprot_id))))
print(paste("Number of duplicated samples: ", sum(duplicated(colnames(mcl[,-c(1:2)])))))
```

### Combine replicates & reshape dataframe from wide to long

I combine the replicated to a dataframe with the average of r1 and r2.

```{r combine replicates}
mcl_r1 <- mcl %>%
  dplyr::select(uniprot_id, contains("r1")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "samples_r1",
                        values_to = "count_R1")

mcl_r2 <- mcl %>%
    dplyr::select(uniprot_id, contains("r2")) %>%
    tidyr::pivot_longer(!uniprot_id,
                        names_to = "samples_r2",
                        values_to = "count_R2")

# mearge mcl_r1 and mcl_r2 and add an average column
mcl_long <- cbind(mcl_r1, mcl_r2[, -1]) # [, -1] to leave out the second uniprot_id column

mcl_avg <- mcl_long %>%
  rowwise() %>%
  mutate(avg_count = mean(c(count_R1, count_R2), na.rm = TRUE)) 

#reformat the NaN in the new column to NA
mcl_avg[mcl_avg == "NaN"] <- NA_integer_
```



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
